---
title: "Analytic Workflow"
author: "Tierney O'Sullivan"
date: "2025-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load library
library(DiagrammeR)
```

# Datasets
See `data/README.md`

# Analytic Workflow

Due to the large data sets and spatial data operations that use a lot of memory, I have broken down the data procurement, cleaning, and analysis into multiple R scripts. See below for a description of each script and their portion of the Analytic workflow. 

### Geographic data

#### Script: create_geographies.R

Uses GADM polygon data to create a subset of polygons to use in this analysis at the ADM0 (country) and ADM1 (state) levels. Uses the LandScan population raster to calculate the population sizes of each ADM1.
```{r, echo=FALSE}
DiagrammeR("
graph TD
      GADM_410.gpkg-- use st_read to import world ADM0 polygons -->GADM0_world
      GADM_410.gpkg-- use st_read to import world ADM1 polygons -->GADM1_world
      DHS_countries_iso.csv-- import relevant country list -->dhs_countries
      LandScan_data-- download world raster pop data 1km -->world_pop_ls
      GADM0_world-- spatial transform to world pop projection -->adm0
      GADM1_world-- spatial transform to world pop projection -->adm1
      adm0-- filter to includ relevant countries -->adm0_subset
      dhs_countries-- filter to includ relevant countries -->adm0_subset
      adm1-- filter to includ relevant countries -->adm1_subset
      dhs_countries-- filter to includ relevant countries -->adm1_subset
      world_pop_ls-- get pop data by summing cells in each polygon -->adm1_pop
      adm1_subset-- get pop data by summing cells in each polygon -->adm1_pop
      adm1_subset-- add pop_size as column to polygon data-->adm1_afr
      adm1_pop-- add pop_size as column to polygon data-->adm1_afr
      adm1_aft-- save data -->ADM1_pop.rds
")
```

### Demographic health surveys (DHS) Diphtheria (DPT3) Vaccine coverage data

#### Script: DHS_API_datadownload.R

File creates a set of surveys that match the countries and years of interest and systematically downloads their survey data for DPT3 vaccine coverage from the DHS API. 
```{r, echo=FALSE}
DiagrammeR("
graph TD
      country_list.csv-- get list of WHO African region countries -->countries
      country_df-- use ISOcodes pkg to get ISO codes -->country_df
      DHS_country_list.csv-- load list of DHS countries surveyed -->dhs_countries
      dhs_countries-- join to add iso codes to dhs countries -->dhs_df
      country_df-- join to add iso codes to dhs countries -->dhs_df
      DHS_survey_list.csv-- load list of all DHS survey ids -->dhs_surveys
      dhs_surveys-- select all relevant surveys -->dhs_surveys
      dhs_surveys-- for loop to download relevant surveys from API -->json_file
      json_data-- get all survey data into a dataframe from list -->json_file
      json_data-- create a massive dataframe of all survey data -->APIdata
      APIdata-- save survey data with metadata in filenames -->DHS_survey_data
      
")
```
#### Script: DHS_geographies.R

Script that creates a spatial data frame with polygons specific to each DHS survey boundary.
```{r, echo=FALSE}
DiagrammeR("
graph TD
      DHS_spatial_boundaries_files.shp-- get filenames of all dhs boundary shapefiles -->file_names
      file_names-- create a spatial dataset of all dhs boundaries -->dhs_spatial
      dhs_spatial-- store spatial file in an r data storage format -->spatial_dhs_boundaries_full_list.RDS
      
")
```
### Conflict data

#### Script: acled_access_key.R

Simple script to store the private API key in a file outside of the main analysis scripts.
```{r, echo=FALSE}
DiagrammeR("
graph TD
      API_key-- save acled access key -->acled_access_key.R
")
```

#### Script: download_acled_api.R

Creates a list of relevant countries and downloads conflict (ACLED) data for each of them via the ACLED API.
```{r, echo=FALSE}
DiagrammeR("
graph TD
      country_list.csv-- get list of WHO African region countries -->countries
      country_df-- use ISOcodes pkg to get ISO codes -->country_df
      acled_iso_codes.csv-- load ACLED country codes from their website -->acled_iso
      acled_iso-- keep ones that are in countries of interest -->acled_iso_use
      country_df-- keep ones that are in countries of interest -->acled_iso_use
      country_df-- join with acled_iso_use -->country_df
      acled_iso_use-- join with country_df -->country_df
      country_df-- filter to exclude countries w missing DHS data -->country_df
      acled_access_key.R-- import acled API key -->acled_api_key
      country_df-- for loop to download 2013-2024 conflict data -->raw_acled_data_files
      acled_api_key-- for loop to download 2013-2024 conflict data -->raw_acled_data_files
      
")
```


### Diphtheria data 
#### Script: download_WHO_weekly_bulletins.R

Creates a script to download all the WHO weekly bulletins for outbreaks and other emergencies from the WHO websites and determines if each contains any information or reporting related to diphtheria outbreaks. This diphtheria status is exported and used for further diphtheria data extraction. 
```{r, echo=FALSE}
DiagrammeR("
graph TD
      WHO_website-- Manually gather WHO Weekly bulletin filenames -->who_weekly_bulletin_filenames.csv
      who_weekly_bulletin_filenames.csv-- load in filename data -->file_names
      file_names-- extract urls for each bulletin -->urls
      urls-- create pdf names to store all bulletins in systematically -->pdf_names
      urls-- use walk2 to download all files -->raw_pdfs
      pdf_names-- use walk2 to download all files -->raw_pdfs
      raw_pdfs-- use map and pdf_text to extract all text -->raw_text
      pdf_names-- use map and pdf_text to extract all text -->raw_text
      raw_text-- use str_detect to find pdfs with diphtheria -->con_diphtheria
      cont_diphtheria-- use lapply and any to get T or F for entire pdf -->cont_diphtheria_ls
      cont_diphtheria_ls-- unlist to create a vector of TF -->cont_diphtheria_vec
      cont_diphtheria_vec-- create df with epiweek file metadata and diphtheria presence absence -->cont_diphtheria_df
      cont_diphtheria_df-- manually examine diphtheria containing pdfs and extract table contents for diphtheria outbreaks -->diphtheria_cases_time_series.csv
      
")
```

### Creating data set for analysis
#### Script: data_collect_diphtheria.R
Combines the data from all sources into two final data sets: a static data set and a time series data set. 

First: geography data
```{r, echo=FALSE}
DiagrammeR("
graph TD
      Exclusion_inclusion_criteria.csv-- get list of WHO African region countries included in analysis & get GID_0 -->country_list
      ADM1_pop.rds-- load geographic data -->polys
      polys-- remove unnecessary countries -->polys
      polys-- make geometries valid -->polys
      polys-- calculate area -->area
      polys-- calculate centroids -->centroids
")
```

Next: Conflict data
```{r, echo=FALSE}
DiagrammeR("
graph TD
      raw_conflict_files-- get names of all acled data -->file_names_acled
      file_names_acled-- read in acled data -->acled_total
      acled_total-- add week start to data -->acled
      acled-- turn into a spatial data set using lat lon -->acled_sf
      acled_sf-- buffer conflict events by 10 km to make polygons -->acled_buff
      acled_buff-- spatial overlay to include acled data within ADM1 -->polys_acled_join
      polys_sf-- spatial overlay to include acled data within ADM1 -->polys_acled_join
      polys_acled_join-- drop geometry and group by week and ADM1 calc weekly conflict -->acled_weekly_data
      acled_weekly_data-- get total conflicts and deaths overall by ADM1 -->acled_total
      polys_idsonly-- create timeseries for all places for all weeks -->acled_ts
      acled_weekly_data-- select relevant cols -->acled_ts2
      rollsum_4yr-- create a function to take rolling sum over 4 yr windown -->rollsum_4yr
      acled_ts-- join to get overall ts dataset -->acled_ts3
      acled_ts2-- join to get overall ts dataset -->acled_ts3
      acled_ts3-- group by ADM1 and get cumulative conflict & fatalities over 4 year moving window -->acled_ts3
      
")
```
Next: DHS time varying data on vaccine coverage
```{r, echo=FALSE}
DiagrammeR("
graph TD
      spatial_dhs_boundaries_full_lust.RDS-- load spatial dhs dataset -->dhs_spatial
      dhs_spatial-- make valid -->dhs_valid
      polys_sf-- spatial overlay intersection -->dhs_spatial_join
      dhs_valid-- spatial overlay intersection -->dhs_spatial_join
      DHS_survey_data-- get names of all DHS survey data -->file_names2
      file_names2-- import all files and bind rows in one df -->dhs_all_files
      dhs_all_files-- select relevant cols -->dhs_surveys
      dhs_spatial_join-- join spatial data and survey data by ADM1 -->dhs_total
      dhs_surveys-- join spatial data and survey data by ADM1 -->dhs_total
      dhs_ts-- create time series of all years and places -->dhs_ts
      dhs_total-- drop geometry and select cols of interest -->dhs_ts2
      dhs_ts-- join with survey data -->dhs_ts3
      dhs_ts2-- join with time series data -->dhs_ts3
      dhs_ts3-- fill missing data with previous estimates -->dhs_ts3
")
```
Next: Diphtheria outbreak status
```{r, echo=FALSE}
DiagrammeR("
graph TD
      diphtheria_cases_time_series.csv-- read in diphtheria data -->diphtheria_df
      rollsum_24wk-- create fun to calc rolling sum over 6 month window -->rollsum_24wk
      diphtheria_df-- create a timeseries of diphtheria cases -->diphtheria_ts
      diphtheria_ts-- create vars cum_cases, new_cases, outbreak_status -->diphtheria_ts
      rollsum_24wk-- create vars new_cases_24wk, outbreak_status -->diphtheria_ts
")
```
Next: Create static dataframe with each ADM1 listed as a row and whether there was ever a diphtheria outbreak reported, and their total cumulative number of conflict events & fatalities
```{r, echo=FALSE}
DiagrammeR("
graph TD
      polys_idsonly-- join -->static
      diphtheria_df-- join -->static
      static-- create outbreak_ever and cum fatalities and conflicts vars -->static
")
```
Finally: Create dataframe of time-varying data with each ADM1 and year-epiweek listed as a row . Create time-varying outbreak status variable, time-varying vaccine coverage variable, and time-varying 4 year cumulative fatalities & conflict events variables.
```{r, echo=FALSE}
DiagrammeR("
graph TD
      timeseries_full-- create full df of all years, isoweeks, ADM1 -->timeseries_full
      dhs_ts3-- join dhs data -->timeseries_full
      acled_ts3-- join acled data -->timeseries_full
      diphtheria_ts-- join diphtheria outbreak data -->timeseries_full
      polys_ids_only-- join for GID_0, COUNTRY, and pop_size vars -->timeseries_full
      time_series_full-- create fatalities per 100k var -->timeseries_full
")
```

### Analytic Workflow for Data Analysis: 
#### Script: FinalReport.Rmd
First: static dataset to get crude relationship between per 100K cumulative conflict-related fatalities vs. whether each ADM1 ever had a diphtheria outbreak from 2017-2024
```{r, echo=FALSE}
DiagrammeR("
graph TD
      static-- create per 100k cum conflict fatality -->static
      static-- create basic glm binomial model to look at crude OR btwn fatal and diphtheria outbreak -->mod
      mod-- generate predictions for range of fatalities -->Predicted_data
      Predicted_data-- plot model predicted prob of diphtheria outbreak -->scatterplot
      mod-- summarize model -->summary_mod
")
```
Second: time-varying dataset to get temporal relationship between per 100K cumulative conflict-related fatalities in the 4 years prior vs. whether each ADM1 has had a diphtheria outbreak in the past 6 months over the entire 2017-2024 time period
```{r, echo=FALSE}
DiagrammeR("
graph TD
      full_timeseries.csv-- load timeseries data-->ts
      ts-- glm accounting for repeated measures and adjusting for vax cov -->mod2
      mod2-- generate predictions for range of fatalities for each ADM1 -->predicted_data_ts
      predicted_data_ts-- plot model predicted prob of diphtheria outbreak -->scatterplot
      mod2-- summarize model -->summary_mod2
")
```

Third: time-varying dataset to get temporal relationship between per 100K cumulative conflict-related fatalities in the 4 years prior vs. whether each ADM1 has had a diphtheria outbreak in the past 6 months over the entire 2017-2024 time period. But now including vaccination as a quadratic term. 
```{r, echo=FALSE}
DiagrammeR("
graph TD
      ts-- create a quadratic variable for vaccination -->ts
      ts-- glm accounting for repeated measures and adjusting for vax cov -->mod3
      mod3-- generate predictions for range of fatalities for each ADM1 -->predicted_data_ts
      predicted_data_ts-- plot model predicted prob of diphtheria outbreak -->scatterplot
      mod3-- summarize model -->summary_mod3
")
```

Fourth: time-varying dataset to get temporal relationship between per 100K cumulative conflict-related fatalities in the 4 years prior vs. whether each ADM1 has had a diphtheria outbreak in the past 6 months over the entire 2017-2024 time period. But now including vaccination as a categorical term. 
```{r, echo=FALSE}
DiagrammeR("
graph TD
      ts-- create categorical vaccination variable -->ts_cat
      ts_cat-- glm accounting for repeated measures and adjusting for vax cov -->mod4
      mod4-- generate predictions for range of fatalities for each ADM1 -->predicted_data_ts
      predicted_data_ts-- plot model predicted prob of diphtheria outbreak -->scatterplot
      mod4-- summarize model -->summary_mod4
")
```
